---
title: "Final Draft"
author: "Maya Casper"
date: "2024-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=F, message=F, warning=F, eval =T) 
```


```{r}
library(pacman)
p_load(haven, tidyverse, modelsummary, here, ggplot2,dplyr, ggtext)
medexam<-read.csv(here('final/MedicalExaminerTableau_2020_05_Migrated Data.csv'))
```

Select out of system type, note, display and number of records 
```{r}
medexam<- medexam %>%
  select(County, Drug.Type,Time.Measure,Count, Time.Period, Rate, Count.rate, Manner)

```

Subsetting to all drugs and statewide data
```{r}
overall<- medexam %>%
  filter(Drug.Type =='All Drugs') %>%
  filter(County =='OREGON STATEWIDE')
```
### Visualization 1 
Oregon, all drugs, all years, only rate
```{r}
overall_rate3<- read.csv(here('final/2019-2023.csv'))
rate<-medexam %>%
  filter(County == "OREGON STATEWIDE") %>%
  filter(Drug.Type =='All Drugs') %>%
  filter(Time.Measure == 'Year') %>%
  filter(Count.rate== 'Rate')


overall_rate <- rate %>%
  group_by(Time.Period) %>%
  summarize(Sum_Rate = sum(Rate))

rate <- rate %>%
  left_join(overall_rate, by = "Time.Period")

rate$Time.Period <- as.numeric(rate$Time.Period)

### 1st rendition 
viz1<- overall_rate3 %>% 
  filter(County == "OREGON STATEWIDE")

ggplot(viz1,
           aes(x = Time.Period, y = Sum_Rate)) +
      geom_line(stat = 'identity') + 
  labs(title = "Drug overdose deaths increasing in Oregon",
           subtitle = 'Especially since 2019',
           x     = '',
           y     = 'Rate per 100,000', 
           caption = "Source: Oregon Prescribing Data Dashboard") + 
      theme(plot.title         = element_markdown(margin = margin(b=10),
                                                  hjust  = 0),
            plot.subtitle      = element_markdown(margin = margin(b=35),
                                                  hjust  = 0, size = 8),
            axis.title.y       = element_markdown(size   = 9, 
                                                  vjust  = 1.025,
                                                  angle  = 360, 
                                                  margin = margin(r = -80,b=-60)),
            panel.background   = element_blank(),
            panel.grid.major.y = element_line(color      = "grey80", 
                                              linewidth  = 0.5, 
                                              linetype   = "dotted"),
            axis.line.x        = element_line(colour='black'),
            axis.text.x        = element_text(margin=margin(t=0)),
            axis.ticks.x       = element_line(),
            axis.ticks.y       = element_blank(),
            axis.ticks.length  = unit(0.25, "cm"),
            plot.caption       = element_text(hjust = 0.01, 
                                              size  = 9,
                                              margin=margin(t=0)), 
            plot.margin = margin(1, 1,1,1, "cm"))+
      scale_y_continuous(breaks = seq(0, 50, by= 5),
                         limits = c(0,50)) +
      scale_x_continuous(expand = c(0,0),
                         limits = c(2009,2022))




```
```
### Visualization 2 
Line graph clustered by county

```{r}
library(dplyr)


counties<-medexam %>%
  filter(Drug.Type =='All Drugs') %>%
  filter(Time.Measure == 'Year') %>%
  filter(Count.rate== 'Rate') 


counties$Time.Period <- as.numeric(counties$Time.Period)

overall_rate2 <- counties %>%
  group_by(County, Time.Period) %>% 
  summarize(Sum_Rate = sum(Rate))
counties <- counties %>%
  left_join(overall_rate2, by = "Time.Period")

overall_rate2<- overall_rate2[-287,]



black<-rep('gray',length(unique(overall_rate3$County)))
    names(black) <- unique(overall_rate3$County) 
    black['OREGON STATEWIDE'] <- 'black' 
  
```


```{r}
ggplot(overall_rate3,
           aes(x = Time.Period, y = Sum_Rate, colour = County)) +
      geom_line(stat = 'identity') + 
      scale_color_manual(values=black)+
      labs(title="There are stark difference in overdose death rates among Oregon counties",
           subtitle = '(*Rates include accidental and suicide overdose deaths*)',
           x     = '',
           y     = 'Rate per 100,000 population',
           caption = "Source: Oregon Prescribing Data Dashboard") +
  theme(plot.title= element_markdown(margin = margin(b=3),hjust=0, size = 12, face ='bold'),
  plot.subtitle= element_markdown(margin = margin(b=10),hjust=0, size = 8),
    panel.background = element_rect(fill='white',colour='white'),
    axis.ticks= element_blank(),
    panel.grid.major.y = element_line(color      = "grey80", 
                                              linewidth  = 0.5, 
                                              linetype   = "dashed"),
    plot.caption= element_text(hjust = 0.01,
                               size = 9,
                               margin=margin(t=0)),
    axis.line.x = element_line(color = "black", linewidth = 0.7),
    axis.ticks.x = element_line(color='black',linewidth = .7), 
    axis.ticks.length=unit(.35, "cm"), 
         plot.margin = margin(1, 1,1,1, "cm"),
         legend.position = 'none', 
         axis.title.y = element_markdown(size   = 7.5, 
                                                  vjust  = 1.001,
                                                  angle  = 360, 
                                                  margin = margin(r = -80,b=0)),) +
      scale_y_continuous(breaks = seq(0, 65, by = 5),
                         limits = c(0,65)) +
      scale_x_continuous(expand = c(0,0),
                         limits = c(2009,2022), 
                         breaks = seq(2010, 2022, by = 2))+ 
      annotate('text',
             x        = 2014,
             y        = 15,
             hjust    = .1,
             vjust    = 0,
             label    = 'Oregon Statewide',
             size     = 3.5,
             color    = 'black') 

#time_series
#install.packages('plotly')
#library(plotly)
#ggplotly(time_series)


```



### By counties
```{r}
overall_rate3<- read.csv(here('final/2019-2023.csv'))


overall_rate4 <- overall_rate3[overall_rate3$County %in% c('GILLIAM', 'LANE', 'MULTNOMAH','BAKER','LINN','KLAMATH'),]

ggplot(overall_rate4,
           aes(x = Time.Period, y = Sum_Rate)) +
      geom_line(stat = 'identity') +
      labs(title="There are stark differences in overdose death rates among Oregon counties",
           subtitle = '(Rates include accidental and suicide overdose deaths)',
           x     = '',
           y     = 'Rate per 100,000',
           caption = "Source:Oregon Prescribing Data Dashboard") +
  theme(plot.title= element_markdown(margin = margin(b=5),hjust=0, size = 12, face ='bold'),
    panel.background = element_rect(fill='white',colour='white'),
    axis.ticks= element_blank(),
    panel.grid.major.y = element_line(color      = "grey95", 
                                              linewidth  = 0.5, 
                                              linetype   = "dashed"),
    plot.caption= element_text(hjust = 0.01, 
                               size = 9,
                               margin=margin(t=0)),
    axis.line.x = element_line(color = "black", linewidth = 0.7),
    axis.ticks.x = element_line(color='black',linewidth = .7), 
    axis.ticks.length=unit(.35, "cm"), 
         plot.margin = margin(1, 1,1,1, "cm") ,
         legend.position = 'none', 
         axis.text.x = element_text(size = 6),
         panel.spacing.x = unit(1, "lines"))+
      scale_y_continuous(breaks = seq(0, 60, by = 5),
                         limits = c(0,60)) +
      scale_x_continuous(expand = c(0,0),
                         limits = c(2009,2022)) + 
      facet_wrap(. ~ County)
```

#drug type
#type<-medexam %>% 
#  filter(Time.Period=='2019') %>% 
#  filter(Time.Measure=='Year') %>% 
#  filter(County=='OREGON STATEWIDE') %>% 
#  filter(Count.rate =='Count') %>% 
#  filter(Manner == 'Accident / Undetermined')



#County 

#highest<-medexam %>% 
 # filter(Time.Period=='2019') %>% 
  #filter(Time.Measure=='Year') %>% 
#  filter(Drug.Type=='All Drugs') %>% 
#  filter(Count.rate== 'Rate') %>% 
#  filter(Manner== 'Accident / Undetermined')






Chloropleth Map
```{r}
library(pacman)
p_load(here, sf, ggplot2, osmdata, tigris, tidyverse, viridis, tidycensus, tmap, elevatr, ggtext)
```


```{r}
map <- st_read(here('/Users/mayacasper/Desktop/EDLD 652/EDLD/final/counties/counties.shp'))

or <- counties(state='41',year=2022)
or$NAME <- toupper(or$NAME)
or <- or %>%
  rename(County = NAME)
  
#Had to change to all caps and rename column 

map <- overall_rate3 %>% 
  filter(Time.Period == 2022) %>%  
filter(County != "OREGON STATEWIDE")

map_plot <- left_join(or, map)


 ggplot() +
  geom_sf(data = map_plot, aes(fill = Sum_Rate), color = "black") + 
     labs(title = 'Multnomah County sees highest rate of overdose deaths in the state in 2022', caption = 'Source: Oregon Prescribing Dashboard', fill = 'Rate per\n 100,000') +
     theme(plot.title= element_markdown(margin = margin(b=5),hjust=.5, size = 12), 
 panel.background = element_rect(fill='white',colour='white'),
    axis.ticks= element_blank(), 
 axis.text.x = element_blank(), 
 axis.text.y = element_blank(), 
 ) +
  scale_fill_viridis()

```



