---
title: "Final Draft"
author: "Maya Casper"
date: "2024-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=F, message=F, warning=F, eval =T) 
```


```{r}
library(pacman)
p_load(haven, tidyverse, modelsummary, here, ggplot2,dplyr, ggtext, here, sf,  osmdata, tigris, viridis, tidycensus, tmap, elevatr, plotly, flexdashboard, ggrepel)

medexam<-read.csv(here('final/MedicalExaminerTableau_2020_05_Migrated Data.csv'))

overall_rate3<- read.csv(here('final/2019-2023.csv'))

map <- st_read(here('/Users/mayacasper/Desktop/EDLD 652/EDLD/final/counties/counties.shp'))
```


```{r}
medexam<- medexam %>%
  select(County, Drug.Type,Time.Measure,Count, Time.Period, Rate, Count.rate, Manner)

```


```{r}
overall<- medexam %>%
  filter(Drug.Type =='All Drugs') %>%
  filter(County =='OREGON STATEWIDE')
```


```{r}

rate<-medexam %>%
  filter(County == "OREGON STATEWIDE") %>%
  filter(Drug.Type =='All Drugs') %>%
  filter(Time.Measure == 'Year') %>%
  filter(Count.rate== 'Rate')


overall_rate <- rate %>%
  group_by(Time.Period) %>%
  summarize(Sum_Rate = sum(Rate))

rate <- rate %>%
  left_join(overall_rate, by = "Time.Period")

rate$Time.Period <- as.numeric(rate$Time.Period)

### 1st rendition 
viz1<-overall_rate3 %>% 
  filter(County == "OREGON STATEWIDE")

ggplot(viz1,
           aes(x = Time.Period, y = Sum_Rate)) +
      geom_line(stat = 'identity') + 
  labs(title = "Statewide overdose deaths continue to rise",
           subtitle = 'Rates double since 2019',
           x     = '',
           y     = 'Rate per 100,000 deaths', 
           caption = "Source: Oregon Prescribing Data Dashboard") + 
      theme(plot.title= element_markdown(margin = margin(b=3),hjust=0, size = 15, face ='bold'),
  plot.subtitle= element_markdown(margin = margin(b=30),hjust=0, size = 12, face='italic'),
            axis.title.y       = element_markdown(size   = 9, 
                                                  vjust  = 1.02,
                                                  angle  = 360, 
                                                  margin = margin(r = -80,b=-80)),
            panel.background   = element_blank(),
            panel.grid.major.y = element_line(color      = "grey80", 
                                              linewidth  = 0.5, 
                                              linetype   = "dotted"),
            axis.line.x        = element_line(colour='black'),
            axis.text.x        = element_text(margin=margin(t=0)),
            axis.ticks.x       = element_line(),
            axis.ticks.y       = element_blank(),
            axis.ticks.length  = unit(0.25, "cm"),
            plot.caption       = element_text(hjust = 0.01, 
                                              size  = 9,
                                              margin=margin(b=5)), 
            plot.margin = margin(1, 1,1,1, "cm"))+
      scale_y_continuous(breaks = seq(0, 50, by= 5),
                         limits = c(0,53)) +
      scale_x_continuous(expand = c(0,0),
                         limits = c(2009,2022))+
  geom_rect(xmin = 2020.3, xmax = 2022, ymin = 0, ymax = 50, fill = "#E15759", alpha = 0.05)+
  geom_segment(aes(x = 2021.2, y = 0, xend = 2021.2, yend = 50), linetype = "dashed", color = "black")+
  annotate('text',
             x        = 2020.65,
             y        = 50.5,
             hjust    = .1,
             vjust    = 0,
             label    = 'COVID-19\n Pandemic',
             size     = 3,
             color    = 'black')+
     annotate("segment", x = 2019, y = 10 , xend = 10, yend = 2021,
              arrow = arrow(type = "closed", length = unit(0.01, "npc")))+
           annotate("segment", x = 2020, y = 10, xend = 2021.1, yend=10,
         arrow = arrow(type = "closed", length = unit(0.01, "npc")))+ 
   annotate('text',
             x        = 2019,
             y        = 10,
             vjust    = 0,
             label    = 'Measure 110 goes\n into effect',
             size     = 3,
             color    = 'black')
  

```
```
### Visualization 2 
Line graph clustered by county

```{r}

counties<-medexam %>%
  filter(Drug.Type =='All Drugs') %>%
  filter(Time.Measure == 'Year') %>%
  filter(Count.rate== 'Rate') 


counties$Time.Period <- as.numeric(counties$Time.Period)

overall_rate2 <- counties %>%
  group_by(County, Time.Period) %>% 
  summarize(Sum_Rate = sum(Rate))
counties <- counties %>%
  left_join(overall_rate2, by = "Time.Period")

overall_rate2<- overall_rate2[-287,]


my_color<-rep('gray',length(unique(overall_rate3$County)))
    names(my_color) <- unique(overall_rate3$County) 
    my_color['OREGON STATEWIDE'] <- 'black' 
    my_color['GILLIAM']<- '#FF9DA7'
    my_color['LANE']<- '#59A14F'
    my_color['MULTNOMAH']<- '#E15759'
    my_color['BAKER']<- '#4E79A7'
    my_color['LINN']<- '#F28E2B'
    my_color['KLAMATH']<- '#9C755F'

finalviz2<- ggplot(overall_rate3,
           aes(x = Time.Period, y = Sum_Rate, colour = County)) +
      geom_line(stat = 'identity') + 
      scale_color_manual(values=my_color) +
      labs(title="There are stark difference in overdose death rates <br /> among Oregon counties",
           subtitle = 'Rates include accidental and suicide overdose deaths',
           x     = '',
           y     = 'Rate per 100,000 deaths',
           caption = "Source: Oregon Prescribing Data Dashboard") +
  theme(plot.title= element_markdown(margin = margin(b=3),hjust=0, size = 15, face ='bold'),
  plot.subtitle= element_markdown(margin = margin(b=30),hjust=0, size = 12, face = 'italic'),
    panel.background = element_rect(fill='white',colour='white'),
    axis.ticks= element_blank(),
    panel.grid.major.y = element_line(color      = "grey80", 
                                              linewidth  = 0.5, 
                                              linetype   = "dashed"),
    plot.caption= element_text(hjust = 0.01,
                               size = 9,
                               margin=margin(t=0)),
    axis.line.x = element_line(color = "black", linewidth = 0.7),
    axis.ticks.x = element_line(color='black',linewidth = .7), 
    axis.ticks.length=unit(.35, "cm"), 
         plot.margin = margin(1, 1,1,1, "cm"),
         legend.position = 'none', 
         axis.title.y = element_markdown(size   = 9, 
                                                  vjust  = 1.05,
                                                  angle  = 360, 
                                                  margin = margin(r = -80,b=0)),) +
      scale_y_continuous(breaks = seq(0, 65, by = 5),
                         limits = c(0,65)) +
      scale_x_continuous(expand = c(0,0),
                         limits = c(2009,2022), 
                         breaks = seq(2010, 2022, by = 2))+ 
      annotate('text',
             x        = 2014,
             y        = 15,
             hjust    = .1,
             vjust    = 0,
             label    = 'Oregon Statewide',
             size     = 4,
             color    = 'black', 
             fontface = 'bold') 
             
ggplotly(finalviz2)

```


```{r Facet Wrap Counties}

overall_rate4 <- overall_rate3[overall_rate3$County %in% c('GILLIAM', 'LANE', 'MULTNOMAH','BAKER','LINN','KLAMATH'),]

finalviz3<- ggplot(overall_rate4,
           aes(x = Time.Period, y = Sum_Rate)) +
      geom_line(stat = 'identity', aes(color = County)) +
      scale_color_manual(values= c('GILLIAM'= '#FF9DA7','LANE'='#59A14F','MULTNOMAH'='#E15759', 'BAKER'= '#4E79A7','LINN' ='#F28E2B', 'KLAMATH' = '#9C755F')) +
      labs(title="There are stark differences in overdose death rates <br />among Oregon counties",
           subtitle = 'Rates include accidental and suicide overdose deaths',
           x     = '',
           y     = 'Rate per <br /> 100,000 deaths',
           caption = "Source:Oregon Prescribing Data Dashboard") +
  theme(plot.title= element_markdown(margin = margin(b=3),hjust=0, size = 12, face ='bold'),
  plot.subtitle= element_markdown(margin = margin(b=30),hjust=0, size = 12, face = 'italic'),
    panel.background = element_rect(fill='white',colour='white'),
    axis.ticks= element_blank(),
    panel.grid.major.y = element_line(color      = "grey95", 
                                              linewidth  = 0.5, 
                                              linetype   = "dashed"),
    plot.caption= element_text(hjust = 0.01, 
                               size = 9,
                               margin=margin(t=0)),
  axis.title.y = element_markdown(size   = 9, 
                                                  vjust  = 1.15,
                                                  angle  = 360, 
                                                  margin = margin(r = -80,b=0)),
    axis.line.x = element_line(color = "black", linewidth = 0.7),
    axis.ticks.x = element_line(color='black',linewidth = .7), 
    axis.ticks.length=unit(.35, "cm"), 
         plot.margin = margin(1, 1,1,1, "cm") ,
         legend.position = 'none', 
         axis.text.x = element_text(size = 8),
         panel.spacing.x = unit(1, "lines"))+
      scale_y_continuous(breaks = seq(0, 60, by = 5),
                         limits = c(0,60)) +
      scale_x_continuous(expand = c(0,0),
                         limits = c(2009,2022)) + 
      facet_wrap(. ~ County)
ggplotly(finalviz3)      
```



```{r Drug Type}
unique(medexam$Drug.Type)

drugtype <- medexam %>% 
  filter(Drug.Type %in% c('Cocaine', 'Fentanyl (including analogues)', 'Heroin', 'Methamphetamine', 'Pharmaceutical Opioid')) %>% 
  filter(County == 'OREGON STATEWIDE') %>% 
  filter(Time.Measure == 'Year') %>% 
  filter(Count.rate == 'Count') %>% 
    filter (Time.Period == '2019')

mycolors<- c('Cocaine' = '#F28E2B', 'Fentanyl (including analogues)' = '#4E79A7', 'Heroin' = '#59A14F', Methamphetamine = '#E15759', 'Pharmaceutical Opioid' = '#FF9DA7') 


drugtype_1 <- drugtype%>%
  group_by(Drug.Type) %>% 
  summarize(Sum_Count = sum(Count))
  
  drugtype_1 <- drugtype_1 %>%
  mutate(new_column = "2019")


ggplot(drugtype_1, aes(x = "", y = Sum_Count, fill = Drug.Type)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  labs(title = "<span style='color:#E15759'>Methamphetamine</span> contributed to the highest number of<br /> fatal overdoses in 2019", x ='', y= '',caption = "Source:Oregon Prescribing Data Dashboard", subtitle= '2019 is the most recently avaliable data')  +
  scale_fill_manual(values=mycolors, name = "")+ 
    theme(plot.title= element_markdown(margin = margin(b=3),hjust=0, size = 15, face ='bold'),
          plot.subtitle= element_markdown(margin = margin(b=30),hjust=0, size = 12, face='italic'),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
    panel.background = element_rect(fill='white',colour='white'),
    axis.ticks= element_blank(),
    plot.caption= element_text(hjust = 0.01, 
    size = 9,
    margin=margin(t=0)))
  

  
```



```{r Chloropleth}
or <- counties(state='41',year=2022)
or$NAME <- toupper(or$NAME)
or <- or %>%
  rename(County = NAME)
  
#Had to change to all caps and rename column 

map <- overall_rate3 %>% 
  filter(Time.Period == 2022) %>%  
filter(County != "OREGON STATEWIDE")

map_plot <- left_join(or, map)

ggplot() +
  geom_sf(data = map_plot, aes(fill = Sum_Rate), color = "black") + 
  labs(title = 'Multnomah County sees highest rate of drug overdose deaths <br /> in the state during 2022', x='', y='', 
       caption = 'Source: Oregon Prescribing Dashboard', 
       fill = 'Rate per<br /> 100,000') +
  theme(plot.title= element_markdown(margin = margin(b=3),hjust=0, size = 15, face ='bold'),
  plot.subtitle= element_markdown(margin = margin(b=30),hjust=0, size = 12, face = 'italic'), 
        legend.title= element_markdown(size = 9),
        panel.background = element_rect(fill = 'white', colour = 'white'),
        axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank())+ 
  annotate('text', 
           y = 45.7, 
           x = -122.3, 
           label = 'Multnomah', 
           size = 2.5) + 
  annotate('text', 
           y = 44, 
           x = -123, 
           label = 'Lane', 
           size = 2.5)  + 
  annotate('text', 
           y = 43.2, 
           x = -117.65, 
           label = 'Malheur', 
           size = 2.5) +
             annotate('text', 
           y = 44.8, 
           x = -120, 
           label = 'Wheeler', 
           size = 2.5) +
  annotate('text', 
           y = 45.45, 
           x = -120.7, 
           label = 'Sherman', 
           size = 2.5, 
           angle = 90) +
  annotate('text', 
           y = 45.4, 
           x = -120.2, 
           label = 'Gilliam', 
           size = 2.5, 
           angle = 90) +
  scale_fill_gradient(low = "#f4f4f4", high ='#E15759')
```


```{r Bar version of chloropheth}
map_bar<-overall_rate3 %>% 
  filter(Time.Period == 2022) %>%
  arrange(desc(Sum_Rate))
  
  ggplot(map_bar, aes(x = reorder(County, Sum_Rate, decreasing=TRUE), y = Sum_Rate)) +
    geom_bar(stat = "identity", fill = '#E15759') + 
    labs(title = '**Multnomah County sees highest rate of overdose deaths in<br /> the state in 2022**',
         x     = '',
         y     = '(Rate per 100,000)', 
         caption = "Source: Oregon Prescribing Data Dashboard") +
   theme(plot.title= element_markdown(margin = margin(b=20),hjust=0, size = 15, face ='bold'),
          axis.title.y = element_text(size = 9, vjust = 1, angle = 360, margin = margin(r = -80)), 
          axis.text.x = element_text(size = 7, angle = 90), 
              panel.background = element_rect(fill='white',colour='white'),
    axis.ticks= element_blank(),
    panel.grid.major.y = element_line(color      = "grey80", 
                                              linewidth  = 0.5, 
                                              linetype   = "dashed"), 
     plot.margin = margin(20, 20, 20, 20, "pt"))
```

```{r}

drugtype2 <- medexam %>% 
  filter(Drug.Type %in% c('Cocaine', 'Fentanyl (including analogues)', 'Heroin', 'Methamphetamine', 'Pharmaceutical Opioid')) %>% 
  filter(County == 'OREGON STATEWIDE') %>% 
  filter(Time.Measure == 'Year') %>% 
  filter(Count.rate == 'Count') %>% 
    filter (Time.Period == '2018')


drugtype_3 <- drugtype2%>%
  group_by(Drug.Type) %>% 
  summarize(Sum_Count = sum(Count))

drugtype_3 <- drugtype_3 %>%
  mutate(new_column = "2018")
  
drugtypefinal<- rbind(drugtype_3, drugtype_1)

drugtypefinal$new_column<-as.character(drugtypefinal$new_column)
  
drugtypefinal$Sum_Count<-as.numeric(drugtypefinal$Sum_Count)


drugtypefinal$label <- ifelse(drugtypefinal$new_column == "2018", 
                               paste(drugtypefinal$Drug.Type, drugtypefinal$Sum_Count, ""), 
                               as.character(drugtypefinal$Sum_Count))

# Set linewidth to 0.5 for all rows
drugtypefinal$Linewidth <- 0.5

# Set linewidth to 1 for rows where new_column is "2018" or "2019"
drugtypefinal$Linewidth[drugtypefinal$new_column %in% c("2018", "2019")] <- 1

# Remove non-numeric characters from the label column
drugtypefinal$label <- gsub("[^0-9.]", "", drugtypefinal$label)

# Plotting
ggplot(drugtypefinal, aes(x = new_column, 
                          y = Sum_Count, 
                          group = Drug.Type, 
                          color = Drug.Type)) +
  geom_line(size = drugtypefinal$Linewidth) +
  geom_point(size = 3, position = position_nudge(y = 0.5)) +
  scale_color_manual(values = mycolors, name = "Drug Type") +
  labs(title = '**Overdose death counts contributed to methamphetamine increased <br />more than any other drug type**',
       x = '',  
       y = '',  
       subtitle = 'Nearly a 30% increase from 2018-2019',
       caption = "Source: Oregon Prescribing Data Dashboard") +
  theme(plot.title = element_markdown(margin = margin(b = 10), hjust = 0, size = 15, face = 'bold'),
        axis.text.x = element_text(size = 7),
        panel.background = element_rect(fill = 'white', colour = 'white'),
        axis.ticks = element_blank(),
        panel.grid.major.y = element_line(color = "grey80", linewidth = 0.5, linetype = "dashed"), 
        plot.margin = margin(20, 20, 20, 20, "pt"), 
        legend.title = element_blank(), 
        legend.text = element_markdown(size = 8), 
        legend.position = 'top') +
  geom_text_repel(aes(label = label),
                  hjust = ifelse(drugtypefinal$new_column == "2018", 1.1, -.9), 
                  vjust = 0.5,
                  size = 3.0, 
                  position = position_nudge(y = 8))
```

