---
title: "Lab 2"
author: "Maya Casper"
date: "2024-02-19"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Section 1

In Section 1, we will focus on analyzing PISA data we used in class. This dataset has math and reading scores from PISA assessment for 10 countries.

To begin working with this dataset, you are required to import and preprocess the data using the following code snippet provided below:

```{r, echo=T,eval=T,comment="",warning=FALSE,message=FALSE}
require(countrycode)
library(here)
library(ggplot2)
library(ggtext)

pisa <- read.csv(here('Lab 2/pisa.csv'))
pisa$iso  <- tolower(codelist[codelist$country.name.en %in% pisa$Country,]$iso2c)
pisa$diff <- pisa$Math - pisa$Reading

```

```{r}
#install.packages("ggflags", repos = c(
 # "https://jimjam-slam.r-universe.dev",
  #"https://cloud.r-project.org"))

library(ggflags)

pisa <- read.csv(here('Lab 2/pisa.csv'))
pisa$iso<- tolower(codelist[codelist$country.name.en %in% pisa$Country,]$iso2c)
pisa$diff <- pisa$Math - pisa$Reading
countlabel<-pisa [1:10,1]


ggplot(pisa, aes(x = reorder(Country, diff, decreasing = FALSE), y = diff)) +
  geom_bar(stat = "identity", width = 0.01, color="black")+
  geom_flag(aes(country = iso), size = 5, y = 1, color = "black") + #### Need to determine the local of flag 
  geom_text(aes(label = Country), hjust = -0.1, vjust = 0.5, size = 3, color = "black") + ## need to move them 
  coord_flip() +
  scale_y_continuous(breaks = seq(-45, 45, by = 10),
    limits = c(-48, 49)) +
  geom_hline(yintercept = 0, color = "black", linetype = "solid")+
  labs(title = '**The difference in mathmatics and reading scores from PISA assessment**',
    x = '',
    y = '') +
  theme(plot.title= element_markdown(margin = margin(b=5),hjust=0, size = 10),
        axis.text.y = element_blank(),
    panel.background = element_rect(fill='white',colour='white'),
    axis.ticks= element_blank(),
    plot.caption= element_text(hjust = 0.01, 
    size = 9,
    margin=margin(t=0)),
    axis.line.x = element_line(color = "black", linewidth = 0.7),
    axis.ticks.x = element_line(color='black',linewidth = .7), 
    axis.ticks.length=unit(.35, "cm"))+
annotate('text',
             x        = 9.2,
             y        = -45,
             hjust    = 0.1,
             vjust    = 0,
             label    = 'Reading scores are higher 
      than Math scores',
             size     = 2.5,
             color    = 'black')+ 
  annotate('text',
             x        = .8,
             y        = 34,
             hjust    = 0.1,
             vjust    = 0,
             label    = 'Math scores are higher 
  than Reading scores',
             size     = 2.5,
             color    = 'black')+ 
   annotate("segment", x = 10, y = -28, xend = 10, yend=-48,
         arrow = arrow(type = "closed", length = unit(0.01, "npc")))+
           annotate("segment", x = 1.6, y = 31, xend = 1.6, yend=49,
         arrow = arrow(type = "closed", length = unit(0.01, "npc")))


```



## Section 2

In Section 2, our analysis will focus on world population. This dataset has the population for 266 countries from 1960 to 2022. 

To begin, you should import the dataset using the following code snippet. This code will first filter the countries and include only European Union countries with at least 10 million people. It will also compute a % change in population from 1960 to 2022.

```{r, echo=T,eval=T,comment="",warning=FALSE,message=FALSE}
library(pacman)
p_load(ggtext)
library(tidyr)
library(ggplot2)
library(ggtext)
pop     <- read.csv(here('Lab 2/population.csv'))

EU <- c("Austria", "Belgium", "Bulgaria", "Croatia", "Cyprus", 
        "Czechia", "Denmark", "Estonia", "Finland", "France", 
        "Germany", "Greece", "Hungary", "Ireland", "Italy", "Latvia", 
        "Lithuania", "Luxembourg", "Malta", "Netherlands", "Poland",
        "Portugal", "Romania", "Slovak Republic", "Slovenia", "Spain","Sweden")

pop <- pop[pop$Country.Name %in% EU,]
pop <- pop[which(pop$X2022>10000000),]
pop$per_change <- ((pop$X2022 - pop$X1960)/pop$X2022)*100


colnames(pop) <- gsub('X', '', colnames(pop))

pop_full <- pop %>% 
  pivot_longer(cols      = starts_with(c('1','2')),
                  names_to  = 'Year',
                  values_to = 'Population') 

pop_full$Year <- as.numeric(pop_full$Year)


my_color        <- rep('gray',length(unique(pop_full$Country.Name))) #creates everything grey 
    names(my_color) <- unique(pop_full$Country.Name)
    my_color["Spain"] <- 'blue'
    my_color["Netherlands"] <- 'orange'
```


```{r, echo=T,eval=F,comment="",warning=FALSE,message=FALSE}
plot<- ggplot(pop_full,
           aes(x = Year, y = Population, colour = Country.Name)) +
      geom_line(stat = 'identity')+
  scale_color_manual(values=my_color)+
 guides(color='none', size = 'none') +   # Hide the legends for color and size
      labs(title="Spain and Netherlands are the two countries with largest\npopulation growth in European Union (1960 -2022)",
           subtitle = '(Among countries with at least 10 million people)',
           x     = '',
           y     = ' Population (in millions)',
           caption = "Source: OECD Stats \nhttps://stats.oecd.org/") +
  theme(plot.title= element_markdown(margin = margin(b=5),hjust=0, size = 12, face ='bold'),
    panel.background = element_rect(fill='white',colour='white'),
    axis.ticks= element_blank(),
    panel.grid.major.y = element_line(color      = "grey80", 
                                              linewidth  = 0.5, 
                                              linetype   = "dashed"),
    plot.caption= element_text(hjust = 0.01, 
                               size = 9,
                               margin=margin(t=0)),
    axis.line.x = element_line(color = "black", linewidth = 0.7),
    axis.ticks.x = element_line(color='black',linewidth = .7), 
    axis.ticks.length=unit(.35, "cm"))




print(plot)
```



## Section 3

In Section 3, our analysis will focus on the relationship between online hotel revenue and the number of travel agents over time using a connected scatterplot.  To begin, you should import the dataset using the following code snippet.

```{r, echo=T,eval=T,comment="",warning=FALSE,message=FALSE}


hotel <- read.csv('hotel.csv',fileEncoding="UTF-8-BOM")
hotel$travel_agents <- hotel$travel_agents/1000

  # number of travel agents are in thousands
  # hotel revenue is in billion dollars

hotel
```

```{r}
  library(ggplot2)
library(ggtext)
first   <-   hotel[1:5,1:3]
second<- hotel [5:9, 1:3]
third<-hotel[9:10, 1:3]
forth<-hotel[10:15, 1:3]
label1<-hotel[1:15,1]
  
  ggplot() + 
    geom_point(data=first,aes(x=hotel_revenue,y=travel_agents), size=2, shape=1) +
    geom_path(data=first,aes(x=hotel_revenue,y=travel_agents),color='#b75b37',linewidth=0.75) + 
    geom_point(data=second,aes(x=hotel_revenue,y=travel_agents),size=2, shape=1) +
    geom_path(data=second,aes(x=hotel_revenue,y=travel_agents),color='#0e73b0',linewidth=0.75) + 
    geom_point(data=third,aes(x=hotel_revenue,y=travel_agents),size=2, shape=1) +
    geom_path(data=third,aes(x=hotel_revenue,y=travel_agents),color='#57a6ac',linewidth=0.75) + 
    geom_point(data=forth,aes(x=hotel_revenue,y=travel_agents),size=2, shape = 1) +
    geom_path(data=forth,aes(x=hotel_revenue,y=travel_agents),color='#b75b37',linewidth=0.75) +
    geom_text(data = hotel, aes(x = ifelse(label1 == "2008", hotel_revenue + 2, hotel_revenue), y = travel_agents, label = label1), 
            hjust = 0, vjust = -1.5, size = 2, color = "black") +
    scale_x_continuous(
      breaks   = seq(0,180,30),
      limits   = c(0,185),
      expand   = c(0,0),
      labels   = paste0(format(seq(0,180,30),digits = 1),'B')) +
    scale_y_continuous(
      breaks   = seq(0,140,20),
      limits   = c(0,150),
      expand   = c(0,0),
      labels   = paste0(format(seq(0,150,20),digits = 1),'K'))+
    labs(title = "**Online Hotel Revenue vs. Number of Travel Agents**", 
         subtitle = '',
         x     = 'Online Hotel Revenue (USD)',
         y     = 'Number of Travel Agents')+ 
    theme(plot.title         = element_markdown(margin = margin(b=10),
                                                  hjust  = -0.1),
          axis.title.x = element_markdown(size = 8, 
                                          vjust = 1.2, 
                                          hjust = 0),
            axis.title.y       = element_markdown(size   = 8, 
                                                  vjust  = 1.075,
                                                  margin = margin(r =0,b=0)), 
          panel.background   = element_blank(),
          panel.grid.major.y = element_line(color='gray',linetype='dashed',linewidth=0.25),
          panel.grid.major.x = element_line(color='gray',linetype='dashed',linewidth=0.25),
          axis.ticks.x       = element_blank(),
          axis.ticks.y       = element_blank(), 
          axis.line.y        = element_line(colour='lightgray'), 
          axis.text = element_markdown(size= 7, 
                                        vjust = 1))+
    annotate('text',
             x        = 30,
             y        = 120,
             hjust    = .1,
             vjust    = 0,
             label    = 'Between 2000 and 2004, online travel revenue increased 
             while the number of travel agents decreased',
             size     = 2.5,
             color    = '#b75b37') +
    annotate('text',
             x        = 60,
             y        = 95,
             hjust    = 0.1,
             vjust    = 0,
             label    = 'From 2004 to 2008 , online hotel revenues continued to increase 
             while the number of travel agents decreased steadily',
             size     = 2.5,
             color    = '#0e73b0')+
    annotate('text',
             x        = 35,
             y        = 58,
             hjust    = 0.1,
             vjust    = 0,
             label    = 'From 2008 to 2009 , online hotel revenue
             ecreased a little but for the first time 
             ince 2000 along with a sudden decrease 
             in the travel of agents',
             size     = 2.5,
             color    = '#57a6ac')+
     annotate('text',
             x        = 102,
             y        = 85,
             hjust    = 0.1,
             vjust    = 0,
             label    = 'Recession',
             size     = 2.5,
             color    = '#57a6ac')+
    annotate('text',
             x        = 120,
             y        = 52,
             hjust    = 0.1,
             label    = 'Online hotel revenue recovered and continued 
             to increase while the number of travel agents 
             kept decreasing and stablized',
             size     = 2.5,
             color    = '#b75b37')+
  annotate("segment", x = 94.5, y = 86.1, xend = 100, yend=86.1,
         arrow = arrow(type = "closed", length = unit(0.01, "npc")))

    
```







